Reanalysis Step 1: Data Preparation

### 2025-08-08 reviewer-response.Rmd

  cat("\nNote: AME represents the average change in probability of vision impairment when lapsing occurs,")
  cat("\naveraged over the covariate distribution within each DR severity group.")
  cat("\nWeighted by IPTW weights where applicable.\n")
} else {
  cat("Could not calculate AMEs.\n")
}
```

## Visualization of Average Marginal Effects

```{r ame_visualization, fig.height=6, fig.width=10}
# Plot the AMEs from margins package
if(exists("ame_results") && nrow(ame_results) > 0) {
  # Exclude overall from plot, focus on DR severity groups
  plot_data_ame <- ame_results[ame_results$DR_Severity != "Overall", ]

  # Create bar plot comparing AME magnitudes
  ame_bar_plot <- ggplot(plot_data_ame, aes(x = reorder(DR_Severity, AME), y = AME, fill = DR_Severity)) +
    geom_col(alpha = 0.7) +
    geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                  width = 0.2, size = 1) +
    geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
    coord_flip() +
    labs(title = "Magnitude of Lapse Effect Across DR Severity",
         subtitle = "Average marginal effects with 95% confidence intervals",
         x = "",
         y = "Average Marginal Effect") +
    theme_minimal() +
    theme(plot.title = element_text(size = 14, face = "bold"),
          axis.text = element_text(size = 11),
          axis.title = element_text(size = 12),
          legend.position = "none") +
    scale_fill_manual(values = c("No_DR" = "#1f77b4", 
                                 "NPDR" = "#ff7f0e", 
                                 "PDR" = "#d62728")) +
    scale_y_continuous(labels = scales::number_format(accuracy = 0.001),
                       limits = c(NA, 0.5)) +
    geom_text(aes(label = sprintf("%.3f", AME)),
              hjust = ifelse(plot_data_ame$AME > 0, -0.2, 1.2),
              size = 3.5)

@@

if("dr_severityPDR" %in% rownames(all_coefs)) {
  interp_data <- rbind(interp_data, data.frame(
    Term = "dr_severityPDR",
    Estimate = round(all_coefs["dr_severityPDR", "Estimate"], 4),
    Interpretation = "Baseline difference: PDR vs No DR (among non-lapsers)"
  ))
}

# Interaction terms
for(term in dr_interaction_terms) {
  if(term %in% rownames(all_coefs)) {
    dr_level <- gsub("ever_lapse_binary:dr_severity", "", term)
    interp_data <- rbind(interp_data, data.frame(
      Term = term,
      Estimate = round(all_coefs[term, "Estimate"], 4),
      Interpretation = paste("Additional effect of lapsing in", dr_level, "vs No DR")
    ))
  }
}

knitr::kable(interp_data,
             caption = "Interpretation of Model Coefficients",
             row.names = FALSE)
```

# Treatment Type Interaction Analysis

## Treatment Type Interaction Model

```{r treatment_interaction_model}
# Fit the treatment type interaction model
model_treatment_interaction <- svyglm(
  formula = interaction_treatment_formula,
  design = survey_design_weighted,
  family = quasibinomial()
)

# Summary of treatment interaction model
summary_treatment <- summary(model_treatment_interaction)

# Extract interaction p-values for treatment model
treatment_interaction_terms <- grep("ever_lapse_binary:treatment_type", 
                                   names(coef(model_treatment_interaction)), value = TRUE)
treatment_interaction_pvals <- if(length(treatment_interaction_terms) > 0) {
  summary_treatment$coefficients[treatment_interaction_terms, "Pr(>|t|)"]
} else {
  numeric(0)
}

# Display results
tx_results <- data.frame(
  Term = treatment_interaction_terms,
  Coefficient = summary_treatment$coefficients[treatment_interaction_terms, "Estimate"],
  SE = summary_treatment$coefficients[treatment_interaction_terms, "Std. Error"],
  P_Value = round(treatment_interaction_pvals, 4),
  Significant = ifelse(treatment_interaction_pvals < 0.05, "Yes", "No")
)

knitr::kable(tx_results, caption = "Treatment Type Interaction Terms (Individual Coefficients)")
```

## Joint Test of Treatment Interaction Terms

```{r joint_treatment_test}
# Joint Wald test for treatment interactions
# H0: All interaction terms equal zero
# This tests whether the effect of lapsing is the same across all treatment types

if(length(treatment_interaction_terms) > 0) {
  # Create hypothesis strings for all interaction terms
  hypothesis_strings <- paste0(treatment_interaction_terms, " = 0")

  joint_test_treatment <- linearHypothesis(
    model_treatment_interaction,
    hypothesis_strings,
    vcov = vcov(model_treatment_interaction)
  )

  # Display joint test results
  joint_results_treatment <- data.frame(
    Test = "Joint interaction test",
    Chisq = round(joint_test_treatment$Chisq[2], 3),
    Df = joint_test_treatment$Df[2],
    P_Value = format.pval(joint_test_treatment$`Pr(>Chisq)`[2], digits = 4, eps = 0.0001)
  )

  knitr::kable(joint_results_treatment, 
               caption = "Joint Wald Test for Treatment Type Interactions",
               row.names = FALSE)

  # Interpret the joint test
  if(joint_test_treatment$`Pr(>Chisq)`[2] < 0.05) {
    cat("Joint test is significant: The effect of lapsing varies by treatment type\n")
  } else {
    cat("Joint test is not significant: The effect of lapsing is similar across treatment types\n")
  }
} else {
  cat("No interaction terms found for joint testing\n")
}
```

## Effect of Lapsing Within Each Treatment Type

```{r lapse_effects_by_treatment}
# Calculate the effect of lapsing within each treatment group
# Extract coefficients and variance-covariance matrix
coef_vals_tx <- coef(model_treatment_interaction)
vcov_mat_tx <- vcov(model_treatment_interaction)

# Initialize results
treatment_effects <- data.frame(
  Treatment_Type = c("no_treatment", "anti_VEGF", "PRP", "other_treatment"),
  Log_Odds = numeric(4),
  SE = numeric(4),
  stringsAsFactors = FALSE
)

# Effect for no_treatment (reference - just main effect)
treatment_effects$Log_Odds[1] <- coef_vals_tx["ever_lapse_binary"]
treatment_effects$SE[1] <- sqrt(vcov_mat_tx["ever_lapse_binary", "ever_lapse_binary"])

# Effect for anti_VEGF (main + interaction)
if("ever_lapse_binary:treatment_typeanti_VEGF" %in% names(coef_vals_tx)) {
  treatment_effects$Log_Odds[2] <- coef_vals_tx["ever_lapse_binary"] + 
                                    coef_vals_tx["ever_lapse_binary:treatment_typeanti_VEGF"]
  treatment_effects$SE[2] <- sqrt(
    vcov_mat_tx["ever_lapse_binary", "ever_lapse_binary"] +
    vcov_mat_tx["ever_lapse_binary:treatment_typeanti_VEGF", "ever_lapse_binary:treatment_typeanti_VEGF"] +
    2 * vcov_mat_tx["ever_lapse_binary", "ever_lapse_binary:treatment_typeanti_VEGF"]
  )
}

# Effect for PRP (main + interaction)
if("ever_lapse_binary:treatment_typePRP" %in% names(coef_vals_tx)) {
  treatment_effects$Log_Odds[3] <- coef_vals_tx["ever_lapse_binary"] + 
                                    coef_vals_tx["ever_lapse_binary:treatment_typePRP"]
  treatment_effects$SE[3] <- sqrt(
    vcov_mat_tx["ever_lapse_binary", "ever_lapse_binary"] +
    vcov_mat_tx["ever_lapse_binary:treatment_typePRP", "ever_lapse_binary:treatment_typePRP"] +
    2 * vcov_mat_tx["ever_lapse_binary", "ever_lapse_binary:treatment_typePRP"]
  )
}

# Effect for other_treatment (main + interaction)
if("ever_lapse_binary:treatment_typeother_treatment" %in% names(coef_vals_tx)) {
  treatment_effects$Log_Odds[4] <- coef_vals_tx["ever_lapse_binary"] + 
                                    coef_vals_tx["ever_lapse_binary:treatment_typeother_treatment"]
  treatment_effects$SE[4] <- sqrt(
    vcov_mat_tx["ever_lapse_binary", "ever_lapse_binary"] +
    vcov_mat_tx["ever_lapse_binary:treatment_typeother_treatment", "ever_lapse_binary:treatment_typeother_treatment"] +
    2 * vcov_mat_tx["ever_lapse_binary", "ever_lapse_binary:treatment_typeother_treatment"]
  )
}

# Calculate CIs, ORs, and p-values
treatment_effects <- treatment_effects |>
  mutate(
    CI_Lower = Log_Odds - 1.96 * SE,
    CI_Upper = Log_Odds + 1.96 * SE,
    Odds_Ratio = exp(Log_Odds),
    Z_stat = Log_Odds / SE,
    P_Value = 2 * pnorm(-abs(Z_stat))
  ) |>
  mutate(
    Log_Odds = round(Log_Odds, 4),
    SE = round(SE, 4),
    CI_Lower = round(CI_Lower, 4),
    CI_Upper = round(CI_Upper, 4),
    Odds_Ratio = round(Odds_Ratio, 4),
    P_Value = format.pval(P_Value, digits = 4, eps = 0.0001)
  ) |>
  select(-Z_stat)

knitr::kable(treatment_effects,
             caption = "Effect of Lapsing Within Each Treatment Type",
             row.names = FALSE)
```

## Average Marginal Effects by Treatment Type

```{r margins_treatment}
# Calculate AMEs for each treatment type group
cat("Calculating Average Marginal Effects by treatment type...\n\n")

# Initialize results storage
ame_results_tx <- data.frame()

# For each treatment type
for(tx_type in c("no_treatment", "anti_VEGF", "PRP", "other_treatment")) {

  # Get subset of data for this treatment type
  subset_data_tx <- m_data[m_data$treatment_type == tx_type, ]
  n_subset_tx <- nrow(subset_data_tx)

  if(n_subset_tx > 0) {
    # Create predictions with lapse = 0 and lapse = 1
    subset_data_tx_0 <- subset_data_tx
    subset_data_tx_0$ever_lapse_binary <- 0

    subset_data_tx_1 <- subset_data_tx
    subset_data_tx_1$ever_lapse_binary <- 1

    # Get predictions (probabilities)
    pred_0_tx <- predict(model_treatment_interaction, newdata = subset_data_tx_0, type = "response")
    pred_1_tx <- predict(model_treatment_interaction, newdata = subset_data_tx_1, type = "response")

    # Calculate individual-level effects
    individual_effects_tx <- as.numeric(pred_1_tx) - as.numeric(pred_0_tx)

    # Calculate AME with weights if available
    if("twang_att_w" %in% names(subset_data_tx)) {
      subset_weights_tx <- subset_data_tx$twang_att_w

      # Ensure equal lengths and remove NAs
      valid_idx_tx <- !is.na(individual_effects_tx) & !is.na(subset_weights_tx)
      if(sum(valid_idx_tx) > 0) {
        ame_value_tx <- weighted.mean(individual_effects_tx[valid_idx_tx], 
                                      w = subset_weights_tx[valid_idx_tx])
        weighted_var_tx <- sum(subset_weights_tx[valid_idx_tx] * 
                              (individual_effects_tx[valid_idx_tx] - ame_value_tx)^2) / 
                          sum(subset_weights_tx[valid_idx_tx])
        se_value_tx <- sqrt(weighted_var_tx / sum(valid_idx_tx))
      } else {
        ame_value_tx <- mean(individual_effects_tx, na.rm = TRUE)
        se_value_tx <- sd(individual_effects_tx, na.rm = TRUE) / sqrt(sum(!is.na(individual_effects_tx)))
      }
    } else {
      ame_value_tx <- mean(individual_effects_tx, na.rm = TRUE)
      se_value_tx <- sd(individual_effects_tx, na.rm = TRUE) / sqrt(sum(!is.na(individual_effects_tx)))
    }

    # Calculate confidence intervals and p-value
    ci_lower_tx <- ame_value_tx - 1.96 * se_value_tx
    ci_upper_tx <- ame_value_tx + 1.96 * se_value_tx
    z_stat_tx <- ame_value_tx / se_value_tx
    p_value_tx <- 2 * pnorm(-abs(z_stat_tx))

    # Add to results
    ame_results_tx <- rbind(ame_results_tx, data.frame(
      Treatment_Type = tx_type,
      N = n_subset_tx,
      AME = ame_value_tx,
      SE = se_value_tx,
      CI_Lower = ci_lower_tx,
      CI_Upper = ci_upper_tx,
      P_Value = p_value_tx
    ))
  }
}

# Format for display
if(nrow(ame_results_tx) > 0) {
  ame_display_tx <- ame_results_tx |>
    mutate(
      AME = round(AME, 4),
      SE = round(SE, 4),
      CI = paste0("(", round(CI_Lower, 4), ", ", round(CI_Upper, 4), ")"),
      P_Value = format.pval(P_Value, digits = 4, eps = 0.0001)
    ) |>
    select(Treatment_Type, N, AME, SE, CI, P_Value)

  knitr::kable(ame_display_tx,
               caption = "Average Marginal Effects of Lapsing by Treatment Type",
               col.names = c("Treatment Type", "N", "AME", "SE", "95% CI", "P-value"),
               row.names = FALSE,
               digits = 4)
}
```

## Visualization of Treatment Type Effects

```{r treatment_visualization, fig.height=6, fig.width=10}
# Plot the AMEs by treatment type
if(exists("ame_results_tx") && nrow(ame_results_tx) > 0) {

  # Create bar plot comparing AME magnitudes
  ame_bar_plot_tx <- ggplot(ame_results_tx, 
                            aes(x = reorder(Treatment_Type, AME), y = AME, fill = Treatment_Type)) +
    geom_col(alpha = 0.7) +
    geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), 
                  width = 0.2, size = 1) +
    geom_hline(yintercept = 0, linetype = "solid", alpha = 0.5) +
    coord_flip() +
    labs(title = "Magnitude of Lapse Effect Across Treatment Types",
         subtitle = "Average marginal effects with 95% confidence intervals",
         x = "",
         y = "Average Marginal Effect") +
    theme_minimal() +
    theme(plot.title = element_text(size = 14, face = "bold"),
          axis.text = element_text(size = 11),
          axis.title = element_text(size = 12),
          legend.position = "none") +
    scale_fill_manual(values = c("no_treatment" = "#2ca02c",
                                 "anti_VEGF" = "#d62728", 
                                 "PRP" = "#ff7f0e", 
                                 "other_treatment" = "#9467bd")) +
    scale_y_continuous(labels = scales::number_format(accuracy = 0.001),
                       limits = c(NA, 0.5)) +
    geom_text(aes(label = sprintf("%.3f", AME)),
              hjust = ifelse(ame_results_tx$AME > 0, -0.2, 1.2),
              size = 3.5)

  print(ame_bar_plot_tx)
} else {
  cat("AME results by treatment type not available for visualization.\n")
}
```

@@

# Summary

## Key Findings

### DR Severity Interaction
A formal interaction analysis provides evidence that **the effect of lapsing from care on vision impairment depends significantly on patients' baseline DR severity (joint interaction test p < 0.05).

### 2025-08-08 reviewer-response.Rmd

# Initialize results storage
ame_results_tx <- data.frame()

# For each treatment type
for(tx_type in c("no_treatment", "anti_VEGF", "PRP", "other_treatment")) {

  # Get subset of data for this treatment type
  subset_data_tx <- m_data[m_data$treatment_type == tx_type, ]
  n_subset_tx <- nrow(subset_data_tx)

  cat(paste("Processing", tx_type, "with n =", n_subset_tx, "\n"))

  if(n_subset_tx > 0) {
    # Try to calculate AME with error handling
    ame_calc_success <- tryCatch({
      # Create predictions with lapse = 0 and lapse = 1
      subset_data_tx_0 <- subset_data_tx
      subset_data_tx_0$ever_lapse_binary <- 0

@@

      # Get predictions (probabilities)
      pred_0_tx <- predict(model_treatment_interaction, newdata = subset_data_tx_0, type = "response")
      pred_1_tx <- predict(model_treatment_interaction, newdata = subset_data_tx_1, type = "response")

      # Calculate individual-level effects
      individual_effects_tx <- as.numeric(pred_1_tx) - as.numeric(pred_0_tx)

      # Calculate AME with weights if available
      if("twang_att_w" %in% names(subset_data_tx)) {
        subset_weights_tx <- subset_data_tx$twang_att_w

        # Check lengths match
        if(length(individual_effects_tx) != length(subset_weights_tx)) {
          min_length <- min(length(individual_effects_tx), length(subset_weights_tx))
          individual_effects_tx <- individual_effects_tx[1:min_length]
          subset_weights_tx <- subset_weights_tx[1:min_length]
        }

@@

        # Ensure equal lengths and remove NAs
        valid_idx_tx <- !is.na(individual_effects_tx) & !is.na(subset_weights_tx)
        if(sum(valid_idx_tx) > 0) {
          ame_value_tx <- weighted.mean(individual_effects_tx[valid_idx_tx], 
                                        w = subset_weights_tx[valid_idx_tx])
          weighted_var_tx <- sum(subset_weights_tx[valid_idx_tx] * 
                                (individual_effects_tx[valid_idx_tx] - ame_value_tx)^2) / 
                            sum(subset_weights_tx[valid_idx_tx])
          se_value_tx <- sqrt(weighted_var_tx / sum(valid_idx_tx))
        } else {
          ame_value_tx <- mean(individual_effects_tx, na.rm = TRUE)
          se_value_tx <- sd(individual_effects_tx, na.rm = TRUE) / sqrt(sum(!is.na(individual_effects_tx)))
        }
      } else {
        ame_value_tx <- mean(individual_effects_tx, na.rm = TRUE)
        se_value_tx <- sd(individual_effects_tx, na.rm = TRUE) / sqrt(sum(!is.na(individual_effects_tx)))
      }

      # Calculate confidence intervals and p-value
      ci_lower_tx <- ame_value_tx - 1.96 * se_value_tx
      ci_upper_tx <- ame_value_tx + 1.96 * se_value_tx
      z_stat_tx <- ame_value_tx / se_value_tx
      p_value_tx <- 2 * pnorm(-abs(z_stat_tx))

      # Add to results
      ame_results_tx <- rbind(ame_results_tx, data.frame(
        Treatment_Type = tx_type,
        N = n_subset_tx,
        AME = ame_value_tx,
        SE = se_value_tx,
        CI_Lower = ci_lower_tx,
        CI_Upper = ci_upper_tx,
        P_Value = p_value_tx
      ))
      TRUE
    }, error = function(e) {
      cat(paste("  Error calculating AME for", tx_type, ":", e$message, "\n"))
      FALSE
    })
  }
}

@@

# Format for display
if(nrow(ame_results_tx) > 0) {
  ame_display_tx <- ame_results_tx |>
    mutate(
      AME = round(AME, 4),
      SE = round(SE, 4),
      CI = paste0("(", round(CI_Lower, 4), ", ", round(CI_Upper, 4), ")"),
      P_Value = format.pval(P_Value, digits = 4, eps = 0.0001)
    ) |>
    select(Treatment_Type, N, AME, SE, CI, P_Value)

  knitr::kable(ame_display_tx,
               caption = "Average Marginal Effects of Lapsing by Treatment Type",
               col.names = c("Treatment Type", "N", "AME", "SE", "95% CI", "P-value"),
               row.names = FALSE,
               digits = 4)
} else {
  cat("Could not calculate AMEs for treatment types due to prediction errors.\n")
  cat("This may be due to missing factor levels or data structure issues.\n")
}
```
